---
title: "Broussonetia RADseq Sample Missingness/MAC Filter and HWE Diagnostics"
subtitle: "Samples retained with <=50% missing genotypes; sites retained with MAC >= 1"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: readable
    df_print: paged
params:
  original_vcf: "VCF/Broussonetia_RADseq.07_tags_refilled.vcf.gz"
  filtered_vcf: "/private/tmp/Broussonetia_RADseq.08_missing50_mac1.vcf.gz"
  filtered_vcf_project: "VCF/filtered/Broussonetia_RADseq.08_missing50_mac1.vcf.gz"
  site_missing25_vcf: "VCF/filtered/Broussonetia_RADseq.09_missing50_mac1_site_missing25.vcf.gz"
  site_missing20_vcf: "VCF/filtered/Broussonetia_RADseq.09_missing50_mac1_site_missing20.vcf.gz"
  metadata_file: "PRJNA437223_Broussonetia_meta.csv"
  filtered_metadata_out: "reports/missing50_mac1/PRJNA437223_Broussonetia_meta.missing50_mac1.csv"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 6,
  dpi = 120
)

local_lib <- normalizePath("R_libs", mustWork = FALSE)
if (dir.exists(local_lib)) {
  .libPaths(c(local_lib, .libPaths()))
}

library(vcfR)
library(tidyverse)
library(DT)
library(patchwork)
library(scales)

theme_set(theme_bw(base_size = 12))
```

# Filtering Step

The filtered VCF was generated by first removing samples with more than 50% missing genotypes, then recalculating INFO tags on the retained samples and removing sites with MAC = 0:

```sh
Rscript scripts/render_missing50_mac1_hwe_report.R

# Inside the render script:
# 1. write reports/missing50_mac1/samples_retained_missing50.txt
# 2. subset samples, recalculate tags, and remove monomorphic sites
bcftools view \
  -S reports/missing50_mac1/samples_retained_missing50.txt \
  -Ou VCF/Broussonetia_RADseq.07_tags_refilled.vcf.gz |
bcftools +fill-tags -Ou -- \
  -t AC,AN,AF,MAF,NS,HWE,ExcHet |
bcftools view -i 'MAF>0' -Oz \
  -o VCF/filtered/Broussonetia_RADseq.08_missing50_mac1.vcf.gz
bcftools index -f VCF/filtered/Broussonetia_RADseq.08_missing50_mac1.vcf.gz

# Optional second-layer site missingness filters from the sample-filtered VCF
bcftools view -i 'F_MISSING<0.25 && MAF>0' -Oz \
  -o VCF/filtered/Broussonetia_RADseq.09_missing50_mac1_site_missing25.vcf.gz \
  VCF/filtered/Broussonetia_RADseq.08_missing50_mac1.vcf.gz
bcftools index -f VCF/filtered/Broussonetia_RADseq.09_missing50_mac1_site_missing25.vcf.gz

bcftools view -i 'F_MISSING<0.20 && MAF>0' -Oz \
  -o VCF/filtered/Broussonetia_RADseq.09_missing50_mac1_site_missing20.vcf.gz \
  VCF/filtered/Broussonetia_RADseq.08_missing50_mac1.vcf.gz
bcftools index -f VCF/filtered/Broussonetia_RADseq.09_missing50_mac1_site_missing20.vcf.gz
```

`MAF > 0` is equivalent to `MAC >= 1` for these biallelic SNPs because a positive minor allele frequency requires at least one observed copy of the minor allele after sample filtering.

# Load Filtered VCF And Metadata

```{r load-data}
original_vcf <- normalizePath(params$original_vcf)
filtered_vcf <- normalizePath(params$filtered_vcf)
filtered_vcf_project <- normalizePath(params$filtered_vcf_project, mustWork = FALSE)
site_missing25_vcf <- normalizePath(params$site_missing25_vcf, mustWork = FALSE)
site_missing20_vcf <- normalizePath(params$site_missing20_vcf, mustWork = FALSE)
metadata_file <- normalizePath(params$metadata_file, mustWork = FALSE)
filtered_metadata_out <- normalizePath(params$filtered_metadata_out, mustWork = FALSE)
report_out_dir <- dirname(filtered_metadata_out)
prefilter_sample_qc_file <- file.path(report_out_dir, "sample_missingness_before_missing50_filter.csv")
removed_samples_file <- file.path(report_out_dir, "samples_removed_missing50.txt")

if (!file.exists(metadata_file)) {
  stop("Metadata file does not exist: ", metadata_file)
}

vcf <- read.vcfR(filtered_vcf, verbose = FALSE)
vcf@gt[, -1] <- gsub("^([0-9.])\\|([0-9.])", "\\1/\\2", vcf@gt[, -1])

sample_ids <- colnames(vcf@gt)[-1]
prefilter_sample_qc <- read.csv(prefilter_sample_qc_file, check.names = FALSE)
removed_samples <- readLines(removed_samples_file, warn = FALSE)
site_missing25_snps <- as.integer(system2("bcftools", c("index", "-n", site_missing25_vcf), stdout = TRUE))
site_missing20_snps <- as.integer(system2("bcftools", c("index", "-n", site_missing20_vcf), stdout = TRUE))

metadata <- read.csv(metadata_file, check.names = FALSE)

if (!"Run" %in% names(metadata)) {
  stop("Metadata file does not contain a Run column: ", metadata_file)
}

for (col in c("Organism", "geo_loc_name_country", "geo_loc_name", "haplotype", "Sample Name")) {
  if (!col %in% names(metadata)) {
    metadata[[col]] <- NA_character_
  }
}

filtered_metadata <- metadata |>
  filter(Run %in% sample_ids) |>
  arrange(match(Run, sample_ids))

missing_metadata <- setdiff(sample_ids, filtered_metadata$Run)
if (length(missing_metadata) > 0) {
  stop(
    "Metadata is missing ",
    length(missing_metadata),
    " filtered VCF sample(s): ",
    paste(head(missing_metadata, 10), collapse = ", "),
    if (length(missing_metadata) > 10) ", ..."
  )
}

stopifnot(identical(filtered_metadata$Run, sample_ids))
write.csv(filtered_metadata, filtered_metadata_out, row.names = FALSE)

summary_tbl <- tibble(
  dataset = c(
    "Original VCF",
    "Samples removed",
    "Sample <=50% missing + MAC >=1",
    "Then site <25% missing",
    "Then site <20% missing",
    "Filtered metadata"
  ),
  samples = c(
    nrow(prefilter_sample_qc),
    length(removed_samples),
    length(sample_ids),
    length(sample_ids),
    length(sample_ids),
    nrow(filtered_metadata)
  ),
  snps = c(
    NA_integer_,
    NA_integer_,
    nrow(vcf@fix),
    site_missing25_snps,
    site_missing20_snps,
    NA_integer_
  )
)

datatable(summary_tbl, rownames = FALSE, options = list(dom = "t"))
```

After filtering, `r length(sample_ids)` samples and `r comma(nrow(vcf@fix))` SNPs remain. The metadata was loaded from `r metadata_file` and the filtered metadata file was written to:

`r filtered_metadata_out`

The project copy of the filtered VCF is:

`r filtered_vcf_project`

The site missingness filtered VCFs are:

`r site_missing25_vcf`

`r site_missing20_vcf`

The report reads the temporary validated copy at `r filtered_vcf` to avoid intermittent OneDrive read timeouts.

The sample missingness filter removed `r length(removed_samples)` samples with missingness >50%. Those sample IDs are written to:

`r removed_samples_file`

```{r metadata-summary}
metadata_summary <- filtered_metadata |>
  count(geo_loc_name_country, Organism, sort = TRUE)

datatable(metadata_summary, rownames = FALSE, options = list(pageLength = 12))
```

# Missing Data By Sample

```{r sample-missingness}
gt <- extract.gt(vcf, element = "GT", as.numeric = FALSE)
called <- !is.na(gt) & gt != "./." & gt != ".|." & gt != "."
sample_missing <- colMeans(!called)
sample_called <- colSums(called)

sample_qc <- tibble(
  Run = names(sample_missing),
  missing_rate = as.numeric(sample_missing),
  called_genotypes = as.integer(sample_called)
) |>
  left_join(
    filtered_metadata |>
      select(Run, Organism, geo_loc_name_country, geo_loc_name, haplotype, `Sample Name`),
    by = "Run"
  ) |>
  arrange(desc(missing_rate))

write.csv(
  sample_qc,
  file.path(report_out_dir, "sample_missingness_after_missing50_mac1.csv"),
  row.names = FALSE
)

datatable(sample_qc, rownames = FALSE, options = list(pageLength = 15))
```

```{r sample-missingness-plots}
p_sample_rank <- sample_qc |>
  mutate(Run = fct_reorder(Run, missing_rate)) |>
  ggplot(aes(Run, missing_rate, color = geo_loc_name_country)) +
  geom_point(size = 2.1, alpha = 0.85) +
  geom_hline(yintercept = c(0.1, 0.2, 0.3, 0.4, 0.5), linetype = "dashed", color = "grey60") +
  coord_flip() +
  scale_y_continuous(labels = percent) +
  labs(x = NULL, y = "Missing genotypes after site filtering", color = "Country")

p_sample_hist <- sample_qc |>
  ggplot(aes(missing_rate)) +
  geom_histogram(bins = 40, fill = "#4C78A8", color = "white") +
  scale_x_continuous(labels = percent) +
  labs(x = "Missing genotypes after site filtering", y = "Samples")

p_sample_rank / p_sample_hist + plot_layout(heights = c(2, 1))
```

```{r sample-missingness-thresholds}
sample_thresholds <- tibble(max_sample_missing = c(0.1, 0.2, 0.3, 0.4, 0.5))

sample_threshold_summary <- sample_thresholds |>
  mutate(
    samples_retained = map_int(max_sample_missing, ~ sum(sample_qc$missing_rate <= .x)),
    samples_removed = length(sample_ids) - samples_retained
  )

datatable(sample_threshold_summary, rownames = FALSE, options = list(dom = "t"))
```

# Hardy-Weinberg Diagnostics

For each SNP, this report compares reference allele frequency against the observed heterozygote genotype frequency. A simple Hardy-Weinberg expectation curve, `2pq`, is added as a reference. Because this dataset spans multiple geographic and taxonomic groups, global HWE filtering should be treated cautiously: population structure can create deviations from HWE even when genotypes are technically sound.

```{r hwe-data}
gt_num <- gt
gt_num[gt_num %in% c("0/0")] <- "0"
gt_num[gt_num %in% c("0/1", "1/0")] <- "1"
gt_num[gt_num %in% c("1/1")] <- "2"
gt_numeric <- matrix(as.numeric(gt_num), nrow = nrow(gt_num), dimnames = dimnames(gt))

n_called <- rowSums(called)
n_het <- rowSums(gt == "0/1" | gt == "1/0", na.rm = TRUE)
alt_allele_count <- rowSums(gt_numeric, na.rm = TRUE)
allele_number <- 2 * n_called
alt_freq <- alt_allele_count / allele_number
ref_freq <- 1 - alt_freq
het_freq <- n_het / n_called
mac <- pmin(alt_allele_count, allele_number - alt_allele_count)

info <- vcfR2tidy(vcf, info_only = TRUE)$fix |>
  as_tibble() |>
  mutate(row_id = row_number())

hwe_df <- tibble(
  row_id = seq_len(nrow(vcf@fix)),
  chrom = vcf@fix[, "CHROM"],
  pos = as.integer(vcf@fix[, "POS"]),
  ref = vcf@fix[, "REF"],
  alt = vcf@fix[, "ALT"],
  n_called = n_called,
  n_het = n_het,
  mac = mac,
  ref_freq = ref_freq,
  alt_freq = alt_freq,
  het_freq = het_freq,
  expected_het_freq = 2 * ref_freq * (1 - ref_freq)
) |>
  left_join(info |> select(row_id, HWE, ExcHet), by = "row_id") |>
  mutate(
    HWE = as.numeric(HWE),
    ExcHet = as.numeric(ExcHet),
    hwe_neglog10 = -log10(pmax(HWE, .Machine$double.xmin))
  )

write.csv(
  hwe_df,
  file.path(report_out_dir, "hwe_site_summary_after_missing50_mac1.csv"),
  row.names = FALSE
)

datatable(
  hwe_df |>
    select(chrom, pos, ref, alt, n_called, mac, ref_freq, het_freq, HWE, ExcHet) |>
    arrange(HWE),
  rownames = FALSE,
  options = list(pageLength = 12)
)
```

```{r hwe-scatter, fig.height=7}
expected_curve <- tibble(ref_freq = seq(0, 1, length.out = 300)) |>
  mutate(expected_het_freq = 2 * ref_freq * (1 - ref_freq))

hwe_df |>
  ggplot(aes(ref_freq, het_freq)) +
  geom_hex(bins = 65) +
  geom_line(
    data = expected_curve,
    aes(ref_freq, expected_het_freq),
    inherit.aes = FALSE,
    color = "#E15759",
    linewidth = 1
  ) +
  scale_fill_viridis_c(option = "C", trans = "sqrt", labels = comma) +
  labs(
    x = "Reference allele frequency",
    y = "Observed heterozygote genotype frequency",
    fill = "SNPs",
    title = "Observed heterozygosity against reference allele frequency",
    subtitle = "Red curve shows Hardy-Weinberg expected heterozygosity, 2pq"
  )
```

```{r hwe-by-pvalue}
hwe_thresholds <- tibble(
  hwe_p_min = c(0, 1e-8, 1e-6, 1e-5, 1e-4, 1e-3, 1e-2, 5e-2)
) |>
  mutate(
    threshold_label = if_else(hwe_p_min == 0, "none", paste0("HWE p >= ", format(hwe_p_min, scientific = TRUE))),
    snps_retained = map_int(hwe_p_min, ~ sum(hwe_df$HWE >= .x, na.rm = TRUE)),
    snps_removed = nrow(hwe_df) - snps_retained,
    percent_retained = snps_retained / nrow(hwe_df)
  )

datatable(hwe_thresholds, rownames = FALSE, options = list(dom = "t"))
```

```{r hwe-threshold-plot}
hwe_thresholds |>
  mutate(threshold_label = factor(threshold_label, levels = threshold_label)) |>
  ggplot(aes(threshold_label, snps_retained)) +
  geom_col(fill = "#4C78A8") +
  geom_text(aes(label = comma(snps_retained)), vjust = -0.25, size = 3.2) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.08))) +
  labs(
    x = "Hardy-Weinberg filter threshold",
    y = "SNPs retained",
    title = "SNP retention under candidate HWE thresholds"
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

```{r hwe-threshold-overlays, fig.height=8}
plot_thresholds <- c(1e-6, 1e-4, 1e-2)

hwe_overlay <- crossing(
  hwe_df |> select(ref_freq, het_freq, HWE),
  hwe_p_min = plot_thresholds
) |>
  mutate(
    threshold = paste0("HWE p >= ", format(hwe_p_min, scientific = TRUE)),
    passes = HWE >= hwe_p_min
  )

hwe_overlay |>
  ggplot(aes(ref_freq, het_freq, color = passes)) +
  geom_point(alpha = 0.22, size = 0.35) +
  geom_line(
    data = expected_curve,
    aes(ref_freq, expected_het_freq),
    inherit.aes = FALSE,
    color = "black",
    linewidth = 0.8
  ) +
  facet_wrap(~ threshold, ncol = 1) +
  scale_color_manual(values = c(`FALSE` = "#E15759", `TRUE` = "#4C78A8")) +
  labs(
    x = "Reference allele frequency",
    y = "Observed heterozygote genotype frequency",
    color = "Passes threshold",
    title = "Candidate HWE thresholds overlaid on heterozygosity pattern"
  )
```

# Session Info

```{r session-info}
sessionInfo()
```
